---
title: "HarvardX Capstone - MovieLens Project"
author: "Stephen Clarke"
date: "18/11/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Executive Summary

The purpose of this report was to create a model for recommending movies using the large data set provided. The data was split into testing and training sets to perform the analyses.
Through data exploration and visualisation techniques, it was found that the genre of films and year of their release do seem to have an impact on their ratings, but these are fairly nominal relative to effect of the users (film raters), and the movies themselves.
The bias from the users and movies was accounted for and used to improve the RMSE of the models to 0.8653. This was then further improved using regularisation, to achieve an RMSE of 0.8648.



## Method and Analysis

I have broken down our 'Method and Analysis' section into four sub-sections; Data Cleaning, Data Exploration, Insights Gained, and Modelling Approach.



#### **Data Cleaning**

To start, we load the required libraries and import the data set, before breaking it into test and train sets, called 'edx' and 'validation'.

```{r message=FALSE, warning=FALSE}
#Load all potential libraries that may be required
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(dslabs)
data(movielens)
library(data.table)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(ggrepel)
ds_theme_set()
library(Lahman)
library(lattice)
library(e1071)
library(caret)
library(knitr)
library(tidyr)
library(stringr)
memory.limit(56000)
```

```{r message=FALSE}
#Create edx set and validation set
dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

# if using R 3.6 or earlier:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                            title = as.character(title),
                                            genres = as.character(genres))


movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
      semi_join(edx, by = "movieId") %>%
      semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```


We then take a look at the data to see what we are working with. Using the 'head' and 'summary' functions, we can get a good overview of the data set.


```{r}
kable(head(edx))

kable(summary(edx))

```

We find that the data set has 6 different headings; userId, movieId, rating, timestamp, title and genre. Additionally, we find that movies in edx data set have a mean rating of 3.512 out of 5, and the most common rating that users give a film is 4.

We count how many individual users there are in the data set, and how many movies.

```{r}
n_distinct(edx$userId)

n_distinct(edx$movieId)
```

There are 69,878 unique individuals who have rated films. 10,677 different movies have been rated.

We then use the 'any' and 'is.na' functions to check for any missing data from the rows and columns in the edx data set.

```{r}
any(is.na(edx))
```

The result comes back as 'FALSE', meaning that there are no blank spaces and the edx data set should be fully workable.



#### **Data Exploration**

The below graph gives a visual representation of what the summary tells us about the edx data set, in a more aesthetically pleasing format. It also shows a more comprehensive look at the spread of ratings given.

```{r}
Avg_Rating <- mean(edx$rating)
edx %>% ggplot(aes(rating)) +
geom_histogram(col='purple4',bins=10,fill='turquoise4') + 
  scale_x_continuous(breaks = seq(0.5,5,0.5)) +
  geom_vline(xintercept=Avg_Rating, col='red4',linetype='solid') +
xlab("Rating") +
  ylab("No. of Ratings") +
  ggtitle("Frequency of Each Rating")
```

What this graph also tells us, is that the standard of films that tend to get watched and rated, tend to be the better films. Furthermore, users tend to provide more positive ratings than negative ones.


To get more information on the edx data set, we must look at the other factors that could contribute to the ratings of movies, starting with the genres.


```{r message=FALSE}
Genre_by_num_ratings <- edx %>% separate_rows(genres, sep = "\\|") %>%
    group_by(genres) %>%
    summarise(ratings_per_genre = n()) %>%
    arrange(desc(ratings_per_genre))

head(Genre_by_num_ratings, n=10)

edx %>% separate_rows(genres, sep = "\\|") %>%
     group_by(genres) %>%
     summarise(ratings_per_genre = n()) %>%
     arrange(desc(ratings_per_genre)) %>%
     head(n=10) %>%
     ggplot(aes(genres, ratings_per_genre)) + geom_col(aes(fill = genres)) +
  xlab("Genre") +
  ylab("No. of Ratings")+
  ggtitle("No. Ratings per Genre")

```

It is clear that different genres of films have different popularity; e.g. Dramas get watched and rated significantly more than Fantasy Films. However, we also want to know how well they have scored.


```{r}

Genre_by_avg_rating <-edx %>% separate_rows(genres, sep = "\\|") %>%
    group_by(genres) %>%
    summarise(rating_of_genre = mean(rating)) %>%
    arrange(desc(rating_of_genre))

head(Genre_by_avg_rating, n=10)

edx %>% 
     separate_rows(genres, sep = "\\|") %>% 
     select(genres, rating) %>% 
     group_by(genres) %>% 
     filter(genres %in% c("Drama","Comedy","Action","Thriller","Adventure")) %>% 
     ggplot(aes(x = genres, y = rating, col = genres)) +
     geom_boxplot()+
  xlab("Genre") +
  ylab("Spectrum of Ratings Given")+
  ggtitle("Ratings of Top 5 Genres")


```

From looking at the top 5 genres (based on how many times they have been rated), it is clear that Dramas tend to score the highest on average, yet also have the lowest scores given. The other top film genres have very similar medians and interquartile ranges.

We convert our timestamp into an actual date to make our data easier to read. We also can separate our movie titles from their release dates to make our dataset easier to work with.

```{r}

edx <- edx %>% mutate(Date_of_Rating = as.Date(as.POSIXct(timestamp, origin="1970-01-01")))

edx <- edx %>% mutate(title = str_trim(title)) %>%  
     extract(title, c('title', 'Year_Released'), regex = '(.*)\\s\\((\\d+)\\)', convert = TRUE)

```

Having added a Release Year and a Date of Rating to our dataset, we can now get more insights into viewing and rating habits.

```{r message=FALSE}

edx %>% 
     separate_rows(genres, sep = "\\|") %>% 
     select(genres, Year_Released, rating) %>%      
  group_by(genres, Year_Released) %>% 
     filter(genres %in% c("Drama","Comedy","Action","Thriller","Adventure")) %>% 
  summarise(rating_of_genre_by_Year = mean(rating)) %>%
  ggplot(aes(x = Year_Released, y = rating_of_genre_by_Year, col = genres)) +
     geom_line() +
  xlab("Year of Film Release") +
  ylab("Average Film Rating") +
  ggtitle("Film Ratings by Genre and Release Date")


```

The above graph shows how some years have had more highly-rated films than others. Separating this into our top 5 genres, gives a good idea of when these genres were at their best. There is a lot of fluctuation year by year for each genre, but it seems that in the last 20-30 years, ratings have been more harsh, or films have been a lower standard on average.

Below, shows how the ratings provided by viewers have varied since they started to be recorded in 1995.

```{r message=FALSE}

edx <- edx %>% mutate(Year_of_Rating = as.numeric(format(Date_of_Rating,'%Y')))

edx %>% 
     separate_rows(genres, sep = "\\|") %>% 
     select(genres, Year_of_Rating, rating) %>% 
     group_by(genres, Year_of_Rating) %>% 
     filter(genres %in% c("Drama","Comedy","Action","Thriller","Adventure")) %>% 
  summarise(rating_of_genre_by_Year_Rated = mean(rating)) %>%
    ggplot(aes(x = Year_of_Rating, y = rating_of_genre_by_Year_Rated, col = genres)) +
     geom_line() +
  xlab("Date of Film Rating") +
  ylab("Average Film Rating") +
  ggtitle("Film Ratings by Genre Over Time")
```

From the above graph it does seem that the users have become slightly more harsh with their film rating over time, but from the graph, this does not look particularly significant.

Out of interest, I followed this up by looking at which of the most viewed films has scored the most highly among the raters.

```{r message=FALSE}

top10_movies <- edx %>% select(title, rating) %>% group_by(title) %>% summarise(ratings_per_film = n(), rating_of_film = mean(rating)) %>% filter(ratings_per_film > 20000) %>% arrange(desc(rating_of_film)) %>% head(10)

kable(top10_movies)

```

Of all films that have been rated more than 20,000 times, 'Shawshank Redemption' is the best rated film in the data set.



#### **Insights Gained**

Through the Data Exploration section, we have looked at numerous factors that seem to change over time, across genres and across films. Genres and the date that films are released, do appear to have an impact from the above graphs.
The movie itself, and the user who rates the movies, will be the first things we look at when trying to get a predictive RMSE. For example, the Shawshank Redemption will generally get a higher score than the average movie. This means that the films will have their own 'bias'. As well as this, users will also have a bias based on their rating history and preferences.
It is clear that across genres, some films get watched a lot more than other, and this is also true of films. Regularisation is required to take into account the fact that Shawshank Redemption has been rated over 28,000 times, while some films may only have been rated once.



#### **Modelling Approach**

Below is the standard formula for calculating root mean squared error (RMSE), which assesses the effectiveness of a predictive model.

```{r}
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2, na.rm = TRUE))
}
```

We ideally want our RMSE to be below 0.87 to indicate that our model is truly predictive.



##### **Naive Baseline RMSE**

We start our approach to working out the RMSE using the Naive Baseline Model.

The formula for this is:

$$Y_{u,i} = \hat{\mu} + \varepsilon_{u,i}$$

Where $\hat{\mu}$ is the mean and $\varepsilon_{i,u}$ is the independent errors centered at 0.

The rating for all films in the data set is as below:

```{r}
mu <- mean(edx$rating)
mu
```


```{r}

naive_rmse <- RMSE(validation$rating, mu)

naive_rmse

model_results <- data.frame(model = "Naive Baseline Model", RMSE = naive_rmse)

model_results

```

The RMSE of the Naive Baseline Model is 1.06, which is far too high to be an acceptable RMSE value.



##### **Using Movie Rating Bias to Improve RMSE**

Earlier the data was partioned into test and train sets to allow for more in depth modelling. These are called validation and edx respectively.

In the data exploration section, it was clear that some movies are generally better than others and therefore, rated more highly.

We add the term $b_i$ to compensate for this, where $b_i$ is the average rating for a film in this data set - or the bias. The new formula will be: 

$$Y_{u,i} = \mu + b_i + \varepsilon_{u,i}$$

To get $b_i$ we use the least squares estimate.

```{r message=FALSE}

movie_avg <- edx %>%
   group_by(movieId) %>%
   summarise(b_i = mean(rating - mu))

```

```{r}

movie_avg

```

We then use our RMSE formula to see how effective this method is for prediction.

```{r}
predicted_ratings <- mu + validation %>% 
  left_join(movie_avg, by = 'movieId') %>% 
  .$b_i

#.$ means pull() function

movie_rating_bias <- RMSE(validation$rating, predicted_ratings)

```


```{r warning=FALSE}

model_results <- bind_rows(model_results, data_frame(model = "Movie Rating Bias Model", RMSE = movie_rating_bias))

model_results

```

By including a factor for the bias of movie ratings, we have reduced the RMSE down to 0.94. This is much better but must be improved upon further for a more robust model.



##### **Using User Rating Bias to Improve RMSE**

Another factor to consider to improve the RMSE, is the preference of users for particular movies. Therefore, we must add in another term, $b_u$, to compensate for this.

$$Y_{u,i} = \mu + b_i + b_u + \varepsilon_{u,i}$$
Where $b_u$ is the user bias.

```{r message=FALSE}

user_avg <- edx %>% 
  left_join(movie_avg, by = 'movieId') %>% 
  group_by(userId) %>% 
  summarize(b_u = mean(rating - mu - b_i))

```

```{r}

user_avg

```

We then use our RMSE formula to see how effective this method is for prediction.
 
```{r}

predicted_ratings <- validation %>% 
  left_join(movie_avg, by = 'movieId') %>% 
  left_join(user_avg, by = 'userId') %>% 
  mutate(b_i_b_u = mu + b_i + b_u) %>%
  .$b_i_b_u

user_bias <- RMSE(validation$rating, predicted_ratings)

```


```{r}

model_results <- bind_rows(model_results, data_frame(model = "Movie and User Bias Model", RMSE = user_bias))

model_results

```

By including a factor for the bias of users, we have reduced the RMSE down to 0.865. This is below our 0.87, and hence, is an effective method of prediction.



##### **Regularisation**

Our Movie Rating Model has significantly reduced RMSE but we must look at what this means for the Movie Ratings themselves.

```{r}

edx_movie_titles <- edx %>% 
select(movieId, title) %>%
distinct()


validation %>% count(movieId) %>%
left_join(movie_avg) %>%
left_join(edx_movie_titles, by="movieId") %>%
arrange(desc(b_i)) %>%
select(title, b_i, n) %>%
head(n=10)

```

When we put the Movies in order from our Movie Rating Model, we notice that some of the top movies have only been rated once. Therefore, we must include another technique to account for this. We will use Regularisation.

Regularisation allows us to penalise films that have got high scores but low sample sizes.

We use λ as a 'tuning parameter' and use cross-validation to get its value.


```{r message=FALSE}

lambdas <- seq(0, 10, 0.25)

mu <- mean(edx$rating)
just_the_sum <- edx %>% 
  group_by(movieId) %>% 
  summarize(s = sum(rating - mu), n_i = n())


rmses <- sapply(lambdas, function(l){
  predicted_ratings <- validation %>% 
    left_join(just_the_sum, by='movieId') %>% 
    mutate(b_i = s/(n_i+l)) %>%
    mutate(pred = mu + b_i) %>%
    pull(pred)
  return(RMSE(predicted_ratings, validation$rating))
})
```


```{r}

qplot(lambdas, rmses)  
lambdas[which.min(rmses)]

min(rmses)

```

The above graph shows the lambda value at which the RMSE is at its lowest, having regularised for movie bias. Then the RMSE is an improvement on when we use our movie bias model alone.

However, we want to use regularisation on the model that accounts for both movie bias and user bias.


```{r message=FALSE}

lambdas <- seq(0, 10, 0.25)

rmses <- sapply(lambdas, function(l){

  mu <- mean(edx$rating)
  
  b_i <- edx %>% 
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu)/(n()+l))
  
  b_u <- edx %>% 
    left_join(b_i, by="movieId") %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - mu)/(n()+l))

  predicted_ratings <- 
    validation %>% 
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(pred = mu + b_i + b_u) %>%
    pull(pred)
  
return(RMSE(predicted_ratings, validation$rating))
})

```


```{r}

qplot(lambdas, rmses)
lambda <- lambdas[which.min(rmses)]
lambda

```

The above graph shows the lambda value at which the RMSE is at its lowest, having regularised for movie and user bias.



## Results

Our table is completed below, having provided regularisation to our movie and user bias model.

```{r}

model_results <- bind_rows(model_results, data_frame(model = "Regularised Movie and User Bias Model", RMSE = min(rmses)))

model_results

```

We see that we get an RMSE of 0.8648170 which means that our model is very predictive.



## Conclusion

I was able to provide an RMSE of 0.8648, which would mean that this system would be provide a solid prediction of what each user would enjoy. To improve the RMSE number further, the genre and 'year of release' could be taken into account, although this would most likely only provide a slight improvement.
